---
title: "Handcoding a Difference in Differences"
output: html_document
permalink: handcoding-difference-in-differences
tags: [R, linear model, panel model, panel data]
published: false
---

In this post we will discuss how to manually implement a Difference-in-Differences (DiD) estimator in R,
using simulated data.

```{r data}
# reproducible random numbers
set.seed(123)

# untreated and treated independent variable
# for period 0
xutr  <- rnorm(1000, mean=5)
xtr   <- rnorm(1000, mean=1)

# create a data.frame with the dep. var., indep. var., time and id vars
# for period 0
dfutr <- data.frame(time = 0, id=   1:1000, y=xutr+15+rnorm(1000), x=xutr)
dftr  <- data.frame(time = 0, id=1001:2000, y=xtr +9+rnorm(1000), x=xtr )
df0    <- rbind(dfutr, dftr)

# repeat for period 1, different means
xutr  <- rnorm(1000, mean=10)
xtr   <- rnorm(1000, mean=2 )

# repeat for period 1
dfutr <- data.frame(time = 1, id=   1:1000, y=xutr+20+rnorm(1000), x=xutr)
dftr  <- data.frame(time = 1, id=1001:2000, y=xtr +18+rnorm(1000), x=xtr )
df1    <- rbind(dfutr, dftr)

# repeat for period 2, different means again
xutr  <- rnorm(1000, mean=15)
xtr   <- rnorm(1000, mean=3 )

# repeat for period 2, now with different effect on dep. var for treated (id=1001:2000)
dfutr <- data.frame(time = 2, id=   1:1000, y=xutr+25+rnorm(1000), x=xutr)
dftr  <- data.frame(time = 2, id=1001:2000, y=xtr +22+rnorm(1000), x=xtr )
df2    <- rbind(dfutr, dftr)
```

Let have a look if the data came out the way we wanted it to.

```{r descriptives}
# summarise data
summary(df0)
summary(df1)
summary(df2)
```

We now combine the observations into a single data.frame and add some dummy variables required for the estimation.

```{r combine}
# combine data
df <- rbind(df0, df1, df2)

# create a dummy for post-treatment
df$post        <- ifelse(df$time  >= 2, TRUE, FALSE)
df$treated     <- ifelse(df$id <= 1000, TRUE, FALSE)
df$interaction <- df$post*df$treated
```

The below plot illustrates what the common trend assumption does.

```{r plot, message=FALSE}
library(ggplot2)
library(dplyr)

df %>%
  group_by(time, treated) %>%
  summarise(y=mean(y)) %>%
  ggplot(aes(x=time)) + 
    geom_line(aes(y=y, colour=treated)) + 
    geom_abline(intercept=10, slope=10, colour="red", linetype=2) # assumed common trend
```

Compute the DiD estimator.

```{r DiDest}
lm(y ~ x + post + interaction, data=df)
```
