---
layout: post
title: "Handcoded Linear Model"
tags: [R, linear model, lm, ols]
---

It can be very insightful to hand code a linear model.

Load a basic data set.

```{r}
data("iris")
```

Inspect the data set.

```{r}
summary(iris)
```

Assign our variables to objects (in the global environment)

```{r}
y <- iris$Petal.Length
x <- iris$Petal.Width
```

Use the built in command.

```{r}
lm( y ~ x -1 ) # the -1 ensures that we don't have an intercept
```

Now we estimate our beta ourselves.

```{r}
(t(x)%*%x)^-1 %*% t(x)%*%y
```

Now lets estimate with an intercept

```{r}
lm( y ~ x )
```

To hand code this, we need to add a vector of ones (1s).
Note that the single `1` that we are binding to the vector `X` will be repeated until it is the same lenght.

```{r}
XI <- cbind(x, 1) # create a new object XI (I for Intercept)
head(XI) # inspect the first 10 rows
```

We also have to correct a small cheat which we used above. `^-1` is not correct syntax for a matrix inversion. In the above case it works correctly because our `X` matrix is in fact a vector. If we pre-multiply this vector with the transpose of itself, we obtain a scalar.

```{r}
dim( t(X) %*% X )
```

This is the reason that our inversion using `^-1` worked correctly. However, for matrices wider than one column this is not the case.

```{r}
dim( t(XI) %*% XI )
(t(XI)%*%XI)^-1
```

We therefore use a different tool, the Generalised Inverse `ginv()` function from the `MASS` package.

Now we can use this matrix to estimate a model with an intercept.

```{r}
library(MASS) # we only need this ones after starting R
ginv( (t(XI)%*%XI) ) %*% t(XI)%*%y
```

Note that this is programmatically exactly the same the way that the `lm()` funcios this.

We can suppress the automatic constant and include our `XI` variable and we will obtain the same results.

```{r}
lm(Y ~ XI -1 )
```

Note that here we have constructed a univariate (uni**variate**) model, however, from a programmic poitn of view, the hurdles of multivariate modelling have already beeno overcome by including the intercept.

It is therefore very easy to use the same method in a case with two independent variables.

```{r}
x1 <- iris$Petal.Width
x2 <- iris$Sepal.Length
y  <- iris$Petal.Length
```

we start by binding the two independent variables together (with vector of `1`s, since we want an intercept).

```{r}
XI <- cbind(x1, x2, 1)
head(XI)
```

Now we estimate our model

```{r}
# make sure the MASS package is still loaded
ginv( (t(XI)%*%XI) ) %*% t(XI)%*%y
```

And that's all! The leap from univariate to multivariat modelling was truely very small.

Now lets look at which method is faster.

```{r}
system.time(lm( y ~ XI ))
system.time(ginv(t(XI)%*%XI) %*% t(XI)%*%y)
```

So far we have been calculating the inverse for pre-multiplication. The faster way to do this is using the QR decomposition (`solve()`).

```{r}
solve(t(XI)%*%XI) %*% t(XI)%*%y
system.time( solve(t(XI)%*%XI) %*% t(XI)%*%y )
```

EDIT: in [tomorrow's post](/handcoded-lm-function) we use the method we developed here to create an easy to use function.
