---
layout: post
title: "Handcoded LM function"
tags: [R, linear model, lm, ols, function, programming]
---

In [yesterday's post](/handcoded-lm) we developed a method for constructing a multivariate linear model with an intercept.

Today we will turn the collection of loose commands into an integrated and easy to use `function`.

A small recap from yesterday. We start by loading data and assigning our variables to objects

```{r}
data(iris)
x1 <- iris$Petal.Width
x2 <- iris$Sepal.Length
y  <- iris$Petal.Length
```

We now construction our linear model, the fastest way of doing this is using the QR decomposition.

```{r}
XI <- cbind(x1, x2, 1) # tie all the independent variables together
solve(t(XI)%*%XI) %*% t(XI)%*%y # solve for beta
```

In `R` functions are treated as objects, which means that we ascribe our function to a name, the same way we ascribed our variables to a name (e.g. `x1 <- Petal.Length`).

In order to create a function, we use the construction function (a function to construct an object of the type function).
Logically, but perhaps also somewhat confusingly, this constructor function is called `function()`.

Lets start with creating a function, called `squaring`, that takes one numeric argument (parameter) and returns it squared.

```{r}
squaring <- function(a) a^2
squaring(a = 2)
```

We can also use negatives of course.

```{r}
squaring(-3)
```

Note that in the above call we did not even specify that the input (`-3`) had to be assigned to the argument `a`,
this was automatically deduced by `R`.

We can also construction a function that takes two arguments and multiplies them.

```{r}
multiplying <- function(a, b) {
  a * b
}
multiplying(5, -6)
```

Note that because this function was a bit longer, I put the code on a new line.
When doing this, the code has to be enclosed in curly brackets (`{` and `}`).

It is now time to apply these insights to our linear model. The first step is to define our function which takes the arguments `y` and `X`.

```{r}
modelling <- function (y, X) {
  
}
```

We can now simply insert our sollution for beta.

```{r}
modelling <- function (y, X) {
  solve(t(XI)%*%XI) %*% t(XI)%*%y # solve for beta
}
```

We can now estimate our model.

```{r}
modelling(y = y, X = XI)
```

Note that the argument (e.g. `X`) and the assigned object (`XI`) do not have to have the same name.

At this point we still need to manually include our incept using `cbind(x1, x2, 1)`.

We can add a small piece to our code, which will allow the function to take care of this.

```{r}
modelling <- function (y, X, intercept=TRUE) {
  if (intercept) X <- cbind(X, 1)
  solve(t(XI)%*%XI) %*% t(XI)%*%y # solve for beta
}
```

The first piece `intercept=TRUE` creates the argument and sets the default to `TRUE`.
That means that, unless me manually specify `default=FALSE`, an intercept will be included.

The second piece `if (intercept) X <- cbind(X, 1)` does the following.

1. check if `intercept == TRUE`
2. if so, overwrite `X` with `cbind(X, 1)`

We can now use this function as follows.

```{r}
X <- cbind(x1, x2)
modelling(y = y, X = X)
```

That gives us a pretty complete linear model.