---
layout: post
title: "Hand Coding the Linear Model"
tags: [R, hand coding, linear model, lm, ols]
permalink: handcoding-lm
---

> In order to understand statistics, you have to do the calculations yourself!

Warnings such as these are often given in statistics courses and for good reason too!
Doing the work yourself really cements the understanding of statistics.

Yet when it comes to econometrics, the main take-aways from the workshops are primarily in terms of the **syntax** of yet another computer program.

In this series of post titled: **handcoded**, I show how many workhorses of the econometrician's toolbox can be implemented in a very simple manner.
This manual implementation greatly helps in keeping an understanding of what actually happens when we call e.g. `MCMC..` with a few paramters.


The Linear Model
----------------------
Then using the **Ordinary Least Squares** approach to solving a model, we start with the following equation.

$$
1+1
$$

We start by loading a basic data set.

```{r}
data("iris")
```

Inspect the data set.

```{r}
summary(iris)
```

Assign our variables to objects (in the global environment)

```{r}
y <- iris$Petal.Length
x <- iris$Petal.Width
```

Use the built in command.

```{r}
lm( y ~ x -1 ) # the -1 ensures that we don't have an intercept
```

Now we estimate our beta ourselves (`solve`! it's coming, read on...).

```{r}
(t(x)%*%x)^-1 %*% t(x)%*%y
```

Now lets estimate with an intercept

```{r}
lm( y ~ x )
```

To hand code this, we need to add a vector of ones (`1`s).
Note that the single `1` that we are binding to the vector `X` will be repeated until it is the same length.

```{r}
XI <- cbind(x, 1) # create a new object XI (I for Intercept)
head(XI) # inspect the first few rows
```

We also have to correct a small cheat which we used above. `^-1` is not correct syntax for a matrix inversion. In the above case it works correctly because our `X` matrix is in fact a vector. If we pre-multiply this vector with the transpose of itself, we obtain a scalar.

```{r}
dim( t(x) %*% x )
```

This is the reason that our inversion using `^-1` worked correctly. However, for matrices wider than one column this is not the case.

```{r}
dim( t(XI) %*% XI )
```

This `^-1` will invert every **individual** number in the matrix, rather than the matrix as a whole.

```{r}
(t(XI)%*%XI)^-1
```

We want to obtain to obtain the inverse of the matrix, because this will allow us to pre-multiply on both sides, eliminating `XI` on the **Right-Hand Side** (RHS).

We therefore use a different tool, the Generalised Inverse `ginv()` function from the `MASS` package.

Now we can use this matrix to estimate a model with an intercept.

```{r}
library(MASS) # we only need this once after starting R
ginv( (t(XI)%*%XI) ) %*% t(XI)%*%y
```

Note that this is programmatically exactly the same the way that the `lm()` funcios this.

We can suppress the automatic intercept and include our `XI` variable and we will obtain the same results.

```{r}
lm(y ~ XI -1 )
```

We have now constructed a univariate (uni**variate**) model, however, from a programmic poitn of view, the hurdles of multivariate modelling have already beeno overcome by estimating a model with an intercept (making `X` a matrix).

It is therefore very easy to use the same method in a case with two independent variables.

```{r}
x1 <- iris$Petal.Width
x2 <- iris$Sepal.Length
y  <- iris$Petal.Length
```

we start by binding the two independent variables together (with vector of `1`s, since we want an intercept).

```{r}
XI <- cbind(x1, x2, 1)
head(XI)
```

Now we estimate our model

```{r}
# make sure the MASS package is still loaded
ginv( (t(XI)%*%XI) ) %*% t(XI)%*%y
```

And that's all! The leap from univariate to multivariat modelling was truely very small.

There is one last thing to note here.
Using [QR decompostion](https://en.wikipedia.org/wiki/QR_decomposition),
we can obtain the inverse that we need in a computationally much more efficient way.
The function that we use for this is `solve()`.

```{r}
system.time(lm( y ~ XI ))
system.time(ginv(t(XI)%*%XI) %*% t(XI)%*%y)
```

So far we have been calculating the inverse for pre-multiplication. The faster way to do this is using the QR decomposition (`solve()`).

```{r}
solve(t(XI)%*%XI) %*% t(XI)%*%y
system.time( solve(t(XI)%*%XI) %*% t(XI)%*%y )
```

EDIT: in [tomorrow's post](/handcoded-lm-function) we use the method we developed here to create an easy to use function.
