---
layout: post
title: WIOD data sets package
tags: [R, wiod, decompr, gvc]
permalink: wiod
published: true
---

The `wiod` package is now available on [CRAN](http://cran.r-project.org/package=wiod).
The package contains the complete [WIOD](http://www.wiod.org/) data sets, in a format compatible with the [decompr](http://qua.st/decompr) and [gvc](http://qua.st/gvc) package.


Installation
-------------
The package can be installed using:

```{r install}
install.packages('wiod') # case sensitive!
```


Usage
----------
Following installation, the package can be loaded using:

```{r}
library(wiod)
```

Data can be loaded using the the `data()` function, using `wiod` followed by the last two digits of the required year, as the argument, e.g.

```{r}
data(wiod95)
```

For information on using the package, please refer to the help files.

```{r}
help('wiod')
help(package = 'wiod')
```
    
For examples of usage, see the function-specific help pages.


```{r}
help('countries')
help('industries')
help('inter95')
```

In addition to the help files we provide long-form examples in the [vignette](http://cran.r-project.org/web/packages/wiod/vignettes/wiod.html).

```{r}
vignette('wiod')

# or
browseVignettes('wiod')
```


Additional Information
-----------------------
An overview of the changes is available in the NEWS file.

```{r}
news(package='wiod')
```

There is also a blog post with information on my [personal website](http://qua.st/).

http://qua.st/wiod/


Development
-------------
A development version can be installed **at your own peril**, using:

```{r, eval=FALSE}
if (!require('devtools')) install.packages('devtools')
devtools::install_github("bquast/wiod")
```

Development takes place on the GitHub page.

http://github.com/bquast/wiod

Bugs can be filed on the GitHub issues page.

https://github.com/bquast/wiod/issues.
