---
layout: post
title: introducing diagonals
tags: [R, diagonals, matrix, package]
permalink: diagonals
---

A new R package `diagonals` is now available on CRAN. The package implements several tools for dealing with **fat diagonals** on matrices, such as this one:

```{r, echo=FALSE, message=FALSE}
library(diagonals)
fatdiag(12, steps=3)
```

Fat diagonal matrices occur when we combine two dimensions of a data set along one edge of a matrix. For example, trade-flow data in the [decompr](http://cran.r-project.org/package=decompr) and [gvc](http://cran.r-project.org/package=gvc) package have each country-industry combination occur on each edge of the matrix. 

The workhorse function of this package is the `fatdiag` function, which tries behave similarly to the `diag` function from the `base` package, but then with diagonals being **fat**.

We can also use the function to assign values to the diagonal.

```{r}
( m <- matrix(111, nrow=6, ncol=9) )
fatdiag(m, steps=3) <- 5
m
```

As can be seen from the above example, the blocks and matrices do not have to be square.

The diagonal of a matrix can also be extracted.

```{r}
fatdiag(m, steps=3)
```

We can also specify the size of the block in stead of the number of steps.

```{r}
fatdiag(12, size=4)
```

This also gives us flexibility in terms of having non-square blocks (and consequently matrices).

```{r}
fatdiag(12, size=c(3,4) )
```

Installation
--------------
The [diagonals packge](http://cran.r-project.org/?package=diagonals) is now [available on CRAN](/diagonals-cran) and can therefore be installed directly from inside `R` using:

```{r, eval=FALSE}
install.packages("diagonals")
```

Subsequently the package can be loaded using:

```{r, eval=FALSE}
library(diagonals)
```

The above introduction is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/fatdiag.html).
It can be accessed from `R` using:

```{r, eval=FALSE}
vignette("fatdiag")

# or
browseVignettes(package = "diagonals")
```


Development
-------------
[![CRAN Version](http://www.r-pkg.org/badges/version/diagonals)](http://cran.r-project.org/?package=diagonals)
[![Travis-CI Build Status](https://travis-ci.org/bquast/diagonals.svg?branch=master)](https://travis-ci.org/bquast/diagonals) 
[![Coverage Status](https://img.shields.io/coveralls/bquast/diagonals.svg)](https://coveralls.io/r/bquast/diagonals?branch=master)

The development version, to be used **at your peril**, can be installed using:

```{r devel, eval=FALSE}
if (!require('devtools')) install.packages('devtools')
devtools::install_github("bquast/decompr")
```

Development takes place on the [GitHub page](http://github.com/bquast/diagonals),
bugs can be filed on the [issues page](https://github.com/bquast/diagonals/issues).
