---
layout: post
title: "Handcoded Categorical Variables"
tags: [R, linear model, lm, ols]
published: false
---

In last week's posts we discussed [handcoding a linear model](/handcoded-lm) and [writing a convenient function for this](/handcoded-lm-function),
in today's post we will take this a step further by including a categorical variable.

Handling categorical variables in a statistical model is bit different because they generally do not have a direct numerical representation.

A special case of categorical variable is the dichotonomous variable. This categorical variable can take only two forms, e.g. `Male|Female`, `Infected|Not-Infected`, `Pre-Intervention|Post-Intervention`, etc.

In this case we use the follow data set.

```{r}
data("Titanic")
```

The data set is currently stored in a `matrix`, for the purposes of our analysis it is more convenient to store it in a `data.frame` object.
We can use the `as.data.frame()` function to convert `Titanic` to a `data.frame` and them overwrite the old `matrix` format object.

```{r}
Titanic <- as.data.frame(Titanic)
```

We can now inspect the `data.frame`.

```{r}
summary(Titanic)
```

We see the follow

This is therefore a dichotonomous variable.

Dichotonomous variables are represented using a so called [dummy variable](https://en.wikipedia.org/wiki/Dummy_variable_(statistics)).
A dummy variable takes on of two values, it has to be either `0` or `1`. 

```{r}
table(Seatbelts)
```
