---
layout: post
title: introducing diagonals
tags: [R, diagonals, matrix, package]
permalink: diagonals
published: true
---

A new R package `diagonals` is now available on CRAN. The package implements several tools for dealing with **fat diagonals** on matrices.

Fat diagonal matrices occur when we combine two dimensions of a data set along one edge of a matrix. For example, trade-flow data in the [decompr](http://cran.r-project.org/package=decompr) and [gvc](http://cran.r-project.org/package=gvc) package have each country-industry combination occur on each edge of the matrix. 

A fat diagonal matrix looks like this.

```{r}
library(diagonals)
fatdiag(12, steps=3)
```

The workhorse function of this package is the `fatdiag` function, which tries behave similarly to the `diag` function from the `base` package, but then with diagonals being **fat**.

We can also use the function to assign values to the diagonal.

```{r}
( m <- matrix(111, nrow=6, ncol=9) )
fatdiag(m, steps=3) <- 5
m
```

As can be seen from the above example, the blocks and matrices do not have to be square.

The diagonal of a matrix can also be extracted.

```{r}
fatdiag(m, steps=3)
```

We can also specify the size of the block in stead of the number of steps.

```{r}
fatdiag(12, size=4)
```

This also gives us flexibility in terms of having non-square blocks (and consequently matrices).

```{r}
fatdiag(12, size=c(3,4) )
```

The [diagonals packge](http://cran.r-project.org/?package=diagonals) is now [available on CRAN](/diagonals-cran) and can therefore be install directly from inside `R` using:

```{r, eval=FALSE}
install.packages("diagonals")
```

Subsequently the package can be loaded using:

```{r, eval=FALSE}
library(diagonals)
```

The above introduction is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/fatdiag.html).
It can be accessed from `R` using:

```{r, eval=FALSE}
vignette("fatdiag")

# or
browseVignettes(package = "diagonals")
```

For more information on the package and its development please see yesterday's post [diagonals on CRAN](/diagonals-cran).
