---
layout: post
title: "Linear Model and Neural Network"
permalink: linear-model-neural-network
tags: [R, linear model, neural network, machine learning]
---

In this short post I want to quickly demonstrate how the most basic neural network (no hidden layer) gives us the same results as the linear model.

First we need data

```{r data}
data(swiss)
str(swiss)
```

We can now specify a model.

```{r model}
m1 <- formula(Fertility ~ Agriculture + Examination + Education + Catholic + Infant.Mortality)
```

Let's start by estimating a linear model.

```{r lm}
lm1 <- lm(formula=m1, data=swiss)
summary(lm1)
```

Now lets use the built in `nnet` package to estimate a neural network without a hidden layer.

```{r nnet}
library(nnet)
nn1 <- nnet(formula=m1, data=swiss, size=0, skip=TRUE, linout=TRUE)
summary(nn1)
```

Let look at the results next to each other.

```{r results, results='hold'}
lm1
summary(nn1)
```

We can also estimate this model using the `neuralnet` package (this package needs to be installed first using `install.packages("neuralnet")`). This package allows us to nicely visualise the results.

```{r neuralnet}
library(neuralnet)
nn1a <- neuralnet(formula=m1, data=swiss, hidden=0, linear.output=TRUE)
plot(nn1a, rep='best')
```

Of course it only becomes interesting if we include a hidden layer.

```{r hidden}
nn2a <- neuralnet(formula=m1, data=swiss, hidden=2)
plot(nn2a, rep='best')
```
