---
title: "Handcoding a Panel Model"
output: html_document
permalink: handcoding-panel-model
tags:
- R
- linear model
- panel model
- panel data
layout: post
---

```{r prepare, message=FALSE}
# load the PLM library for panel estimation
library(plm)
# load the Crime data set
data(Crime)
```

The most basic panel estimation is the Pooled OLS model,
this model combines all data across indices and performs a regular Ordinary Least Squares Estimation.

```{r pooled-ols}
# define the model
m1 <- formula(crmrte ~ prbarr + prbconv + polpc)

# create a panel data.frame (pdata.frame) object
PanelCrime <- pdata.frame(Crime, index=c("county", "year") )

# estimate Pooled OLS using the basic lm function
lm(formula = m1,
   data    = Crime)

# estimate the Pooled OLS using the plm package
plm(formula = m1,
    data    = PanelCrime,
    model   = "pooling"  )
```

A more complex estimation method is the Fixed-Effect (or within) estimator.
If our data only contains to time-periods,
the results of this estimator are equivalent to a OLS estimation of the first-differenced variables.

```{r first-difference}
# create data.frame with only years 81 and 82
Crime8182      <- subset(Crime, year %in% c(81, 82) )
# put into panel data.frame form (pdata.frame)
PanelCrime8182 <- pdata.frame(Crime8182, index=c("county", "year") )

# first difference the non-panel data.frame
library(dplyr)
Crime8182FD <- Crime8182 %>%
  group_by(county) %>%
  summarise(crmrte  = diff(crmrte),
            prbarr  = diff(prbarr),
            prbconv = diff(prbconv),
            polpc   = diff(polpc)   )

# use lm to estimate the two-period fixed-effects model
lm (formula = m1,
    data    = Crime8182FD    )

# verify with the plm package
plm(formula = m1,
    data    = PanelCrime8182,
    model   = "fd"           )
```

If our data set contains more than two time periods,
we need to estimate an proper fixed effects model.
This is done by creating a fixed-effect variable for every level along the cross-sectional index (i.e. the non-time index).
A simple way of doing this, is by encoding the cross-section index as a factor and including that factor in the regression (more on factors/categorical variables in the post on [Handcoding a Linear Model](/handcoding-linear-model)).

```{r fixed-effect}
fe <- lm (formula = crmrte ~ prbarr + prbconv + polpc + factor(county),
          data    = Crime)
fe$coefficients[2:4]

      plm(formula = m1,
          data    = PanelCrime,
          model   = "within"   )
```
