---
layout: post
title: introducing diagonals
tags: [R, diagonals, matrix]
permalink: diagonals
---

A new R package `diagonals` is now available on CRAN. The package implements several tools for dealing with **fat diagonals** on matrices.

Block-diagonal matrices can be extracted or removed using two small
functions implemented here.
Furthermore, non-square matrices are
supported using two additional functions.

Block-diagonal matrices occur when we combine two dimensions of a data
set along one edge of a matrix. For example, trade-flow data in the
[decompr](http://cran.r-project.org/web/packages/decompr/index.html) and
[gvc](http://cran.r-project.org/web/packages/gvc/index.html) package
have each `country * industry` combination occur on each edge of the matrix.

```{r}
m <- matrix(1:64, nrow=8)
m
```

We can now drop everything except a block diagonal.

```{r}
library(diagonals)
block_matrix(m, steps=4)
```

Or drop only the block diagonal.

```{r}
minus_block_matrix(m, steps=2)
```

The number of `steps` can vary, but dimensions should be multiples of
`steps`. Alternatively we can specify the size of the `size` of the
step.

```{r}
block_matrix(m, size=2)
```

Even non-square matrices are supported, as long as each dimension is a
multiple of size.

```{r}
# create the matrix
nsm <- matrix(1:27, nrow=9, ncol=3 )

# test if the dimensions are multiples of size
# i.e. there are no remainders after the modulo division of dimensions by steps
dim(nsm) %% 3 # we will use 3 (three) steps
```

Now extract the rectangular diagonal.

```{r}
rectangle_matrix(nsm, steps=3)
```

Or this.

```{r}
rectangle_matrix( t(nsm), steps=3 )
```

Lastly, we can also choose a replacement parameter other than zero.

```{r}
block_matrix(m, steps=4, replacement=-1)
```

or

```{r}
minus_rectangle_matrix(nsm, steps=3, replacement=666)
```

The [diagonals packge](http://cran.r-project.org/?package=diagonals) is now [available on CRAN](/diagonals-cran) and can therefore be install directly from inside `R` using:

```{r, eval=FALSE}
install.packages("diagonals")
```

Subsequently the package can be loaded using:

```{r, eval=FALSE}
library(diagonals)
```

The above introduction is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/diagonals.html).
It can be accessed from `R` using:

```{r, eval=FALSE}
vignette("diagonals")
```

An application of `diagonals` in **Social Network Analysis** is demonstrated in last weeks post: [diagonals in networks](/diagonals-network). This post is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/network.html)

For more information on the package and its development please see yesterday's post [diagonals on CRAN](/diagonals-cran).
