---
layout: post
title: "Hand Coding Instumental Vairables"
tags: [R, hand coding, linear model, lm, iv, instumental variables, 2sls, tsls]
permalink: handcoding-instrumental-variables
---

In a previous post we discussed the [linear model](/handcoding-lm)
and how to write a [function that performs a linear regression](/handcoding-lm-function).
In this post we will use that linear model function to perform a [Two-Stage Least Squares estimation].
This estimation allows us to [...]

[Recall](/handcoding-lm-function) that we built the follow linear model function.

```{r}
ols <- function (y, X, intercept=TRUE) {
  if (intercept) X <- cbind(X, 1)
  solve(t(X)%*%X) %*% t(X)%*%y # solve for beta
}
```

We used the following data.

```{r}
data(iris)
x1 <- iris$Petal.Width
x2 <- iris$Sepal.Length
y  <- iris$Petal.Length
```

This allowed us to estimate a linear model.

```{r}
X <- cbind(x1, x2)
ols(y = y, X = X)
```

Which includes the intercept, since the default value it `TRUE` (see function definition above),
we could estimate it without an intercept using.

```{r}
ols(y = y, X = X, intercept = FALSE)
```

Having revisited the above, we can continue with instumental variables.
We will replicate an example from the `AER` (Applied Econometric Regressions) package.

```{r}
library(AER)
data("CigarettesSW")
rprice  <- with(CigarettesSW, price/cpi)
tdiff   <- with(CigarettesSW, (taxs - tax)/cpi)
packs   <- CigarettesSW$packs
```

We first need to obtain our first stage estimate (putting the whole function between parentheses allows us to both write it to the object `s1` and print it.

```{r}
( s1 <- ols(y = rprice, X = tdiff) )
```

We can now obtain the predicted (fitted) values

```{r}
Xhat <- s1[1]*tdiff + s1[2]
```

Using these fitted values, we can finally estimate our second stage.

```{r}
ols(y = packs, X = Xhat)
```

Now compare this to the results using the `AER` package.

```{r}
library(AER)
ivreg(packs ~ rprice | tdiff)
```

When we compare this estimate to the estimate from the linear model, we find that the effect of the price is significantly overestimated there.

```{r}
lm(packs ~ rprice)
```

We can also do this using `R`'s built in `lm` function.

```{r}
# first stage
s1 <- lm(rprice ~ tdiff)

# estimate second stage using fitted values (Xhat)
lm(packs ~ s1$fitted.values)
```

As an intermediate form, we can  manually calculate `Xhat` (`fitted.values`) and estimate using that.

```{r}
# manually obtain fitted values
Xhat <- s1$coefficients[2]*tdiff + s1$coefficients[1]

# estimate second stage using Xhat
lm(packs ~ Xhat)
```

Note that if you estimate a TSLS using the `lm` function,
that you can **only** use the coefficients,
the error terms will be **wrong**. 
