---
layout: post
title: diagonals in a network
tags: [R, diagonals, network, social]
permalink: diagonals-network
---

A typical example in which `diagonals` can be helpful is Social Network
Analysis. For example, if we use matrices to represent friendship
perceptions between individuals, then we need a dyadic matrix.

```{r}
# generate a dyadic matrix for 3 individuals
m <- matrix(sample(0:1, 9, replace=TRUE), nrow=3, ncol=3)
m
```

Let says that we want to look at second-order connections (i.e. friends
of friends). If we now want to represent the data from both time period
in a single object, we need a **4-dimensional array**. Higher-order arrays
are hard to visualise, another way of doing this is by representing two
dimensions along each of the two edges of a matrix. We can do this using
the **Knonecker Product** (denoted ⊗), which we can call in `R` using
the alias `%x%`.

```{r}
M <- m %x% m
M
```

Feelings of friendship towards oneself aren't always particularly
insightful. We can now use the `diagonals` library to eliminate those.

```{r}
# load the library
library(diagonals)

# remove the elements along the diagonal of width 2
minus_block_matrix(M, size=3)
```

The [diagonals](http://cran.r-project.org/?package=diagonals) package now [available on CRAN](/diagonals-cran) and can therefore be install directly from inside `R` using:

```{r, eval=FALSE}
install.packages("diagonals")
```
Subsequently the package can be loaded using:

```{r, eval=FALSE}
library(diagonals)
```

The above demonstration is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/network.html).
It can be accessed from `R` using:

```{r, eval=FALSE}
vignette("network")
```

A general introduction to `diagonals` is available in next weeks post: [diagonals](/diagonals). This post is also available as a vignette that is [included in the package](http://cran.r-project.org/web/packages/diagonals/vignettes/diagonals.html)

For more information on the package and its development please see yesterday's post [diagonals on CRAN](/diagonals-cran).
