---
layout: post
title: "sigmoid package"
tags: [R, neural network, sigmoid]
permalink: sigmoid
---

The [sigmoid package ](http://cran.r-project.org/package=sigmoid) makes it easy to become familiar with the way neural networks work by demonstrating the key concepts using straightforward code examples.

Installation
--------------
The package can now be installed from [CRAN](http://cran.r-project.org/) using:

```{r, eval=FALSE}
install.packages('sigmoid') # case sensitive!
```

Usage
---------
After installation, the package can be loaded using:

```{r}
library(sigmoid)
```

For information on using the package, please refer to the help files.

```{r}
help('sigmoid')
help(package = 'sigmoid')
```

Additional Information
-----------------------
An overview of the changes is available in the NEWS file.

```{r}
news(package='sigmoid')
```


Development
-------------
A development version can be installed **at your own peril**, using:

```{r, eval=FALSE}
if (!require('devtools')) install.packages('devtools')
devtools::install_github("bquast/sigmoid")
```

Development takes place on the [GitHub page](https://github.com/bquast/sigmoid),
bugs can be filed on the [Issues page](https://github.com/bquast/sigmoid/issues).
